---
title: "MANZI SHEMA Christian"
author: "ID101382"
date: "2025-10-16"
output: html_document
---

## R for data science class 1 
### Important definitions in data science

```{r }

```

## Data structure (Vectors, Factors, Lists, Data frame, matrix and arrays)
### Vector: The foundational R data structure

```{r}
# Numeric
x <- c(0.5, 0.6)
age<-c(20,35,32,29)
summary(age)
```
```{r}
age<-c(35,24,18,24)
mean(age)
```
#Logical
```{r}
x <- c(TRUE, FALSE)
x <- c(T, F)
```
## Character

```{r}
x <- c("a", "b", "c" )
class<-c("M","F","F","M")
```
## Integer

```{r}
x <- 9:29
```
## Matrix and operation within matrices

```{r}
x<- matrix(1:6, nrow = 2, ncol = 3)
x
```
```{r}
(y<-matrix(1:6, nrow = 3, ncol = 2))
```
```{r}
x%*%y
```
## List
```{r}
x <- list(1, "a", TRUE, 1 + 4i)
x
```
##Factor
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
```
## Create a vector with NAs in it

```{r}
x <- c(1, 2, NA, 10, 3)
```
## Return a logical vector indicating which elements are NA
```{r}
is.na(x)
```
```{r}
x <- c(1, 2, 4, "NA", 5)
bad <- is.na(x)
print(bad)
```
```{r}
x[!bad]
```
### What if there are multiple R objects and you want to take the subset with no missing values in any of those objects?
```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
```
## Coercion

If character is present, in a vector, R convert everything in the vector to character strings. 

If a vector only contains logical and numbers, R will convert the logical to numbers, Every true becomes a 1, and every FALSE becomes 0
```{r}
sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))
```
## Create data
Create data Using data frame Data frame is more general than matrix How???
Because data frame can contain different modes of data (Numeric, character and so on) Similar to what you can see in SPSS, SAS,…

```{r}
studentID<-c(1,2,3,4,5)
math_score<-c(12,17,10,9,NA)
gender<-c("M","F","M","M","F")
it_score<-c(13,18,11,10,19)
scoredata<-data.frame(studentID,gender,math_score,it_score)
scoredata

#View(scoredata)
```
## Create data from keyboard

Steps 1. Create data frame (or matrix) with variable names 2. Invoke the text editor in the data objected created at first step

```{r}
data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric
(0))
data_class2<-edit(data_class2)
```
## Data Importation

R has some features that can allow to import data from different sources ( It can be text file, spreadsheet, or database)
* 1. Data from excel
* 2. Data statistical packages (SAS, SPSS, Stata)
* 3. Data from Text files (ASCII, XML, Webscraping)
* 4. Data from database management systems (SQL,MySQL, Oracle, Access)
